---
title: "R Notebook"
output: html_notebook
---
Based on Allen 2017 paper
From README on github:
The file data/predictions.RData contains the output of the model. All variables are described in more depth in the paper.

The columns lon and lat are WGS84 longitude and latitude. gridid is a grid cell identifier.
+ bsm_weight_pubs is the model’s output weighted with publications (“observed”).
+ bsm_weight_pop is the model’s output weighted with population (“bias factored out”).
+ bsm_response is the model’s output weighted with neither.
The other columns are generally predictor variables used in the model.

I used bsm_response on the data Peter suggested. But it produced figure 4. 

However, on the metadata of the file I download bsm_response is described as:
https://github.com/ecohealthalliance/hotspots2/blob/a27520e8d6fe1950cb4b88299efcd229e36135b9/inst/out/unbl_data/metadata.csv
bsm_response: The binomial probability that a grid cell sampled at that location will contain an EID event as opposed to a background sample, when drawing equal numbers of absence and background samples weighted by reporting effort.

```{r}
# Repro map of zoonotic EID hotspots (Allen et al. 2017) in R
# Packages
pkgs <- c("terra","sf","rnaturalearth","rnaturalearthdata","tidyterra","ggplot2")
invisible(lapply(setNames(pkgs, pkgs), function(p) if (!requireNamespace(p, quietly=TRUE)) install.packages(p)))
library(terra); library(sf); library(rnaturalearth); library(tidyterra); library(ggplot2)

# 1) Download authors' Zenodo archive with data/figures/code
# (Allen et al. 2017 "hotspots2" v1.0r2)
eid_risk_relative_to_reporting_effor <- read_csv("~/Library/CloudStorage/OneDrive-StockholmUniversity/KVA backup/KVAOneDrive_backup_28Jan2026/2. Projects/4.Cascades/Editorial Review 20250807/map/Anthropocene-maps/6.Infectious_diseases/eid_risk_relative_to_reporting_effor.csv")
data <- eid_risk_relative_to_reporting_effor
```

#Figure 4

```{r}
# Land mask (so country borders are visible)
land_sf     <- ne_countries(scale = "medium", returnclass = "sf") 

ggplot() +
  # geom_polygon(aes(x = lon, y = lat, group = group), data = map.world(), inherit.aes = FALSE, fill = viridis(1)) +
  geom_raster(aes(x = lon, y = lat, fill = bsm_response), data = eid_risk_relative_to_reporting_effor) +
   # geom_path(aes(x = lon, y = lat, group = group), data = map.world(), inherit.aes = FALSE, color = "white", size = 0.15) +
  geom_sf(data =land_sf, fill = NA, color = "white", size = 0.15) +
  #  coord_fixed() +
  ylim(-65, 90) +
  scale_fill_gradientn(colours = viridis(100),
                       guide = guide_colorbar(label = TRUE,
                                              label.position = "right",
                                              title = "Event probability\n(relative to\nreporting effort)")) +
#  theme_black_legend() +
  theme(panel.background = element_rect(fill = "black"),
        plot.background  = element_rect(fill = "black"),
        line = element_blank(),
        legend.title = element_text(color = "white", size = 8),
        legend.text = element_text(color = "white", size = 8),
        legend.title.align = 0,
        legend.background = element_blank(),
        legend.position = c(0.11, 0.45)) +
  labs(x = NULL, y = NULL)
```
