---
title: "Reproducing Figure 3A from Allen et al. (2017)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

if (!requireNamespace("librarian", quietly = TRUE)) {
  install.packages("librarian", repos = "https://cloud.r-project.org")
}

library(librarian)
shelf(raster, terra, sf, sp, rnaturalearth, tidyverse, viridis, maps)
```

# Overview

This document reproduces Figure 3A from Allen et al. (2017), *Global hotspots and correlates of emerging zoonotic diseases*.

Figure 3A is the predicted distribution of observed zoonotic EID event risk, using the weighted model output with current reporting effort.

Source:
[Allen et al. (2017), Nature Communications](https://www.nature.com/articles/s41467-017-00923-8)

# 1. Load archived model predictions

```{r load-predictions}
predictions_path <- "~/Library/CloudStorage/OneDrive-StockholmUniversity/KVA backup/KVAOneDrive_backup_28Jan2026/2. Projects/4.Cascades/Editorial Review 20250807/map/Base maps/6.Infectious_diseases/Allen 2017/predictions.RData"

stopifnot(file.exists(path.expand(predictions_path)))
load(path.expand(predictions_path))  # loads object: predictions

predictions <- predictions %>%
  dplyr::select(gridid, lon, lat, bsm_weight_pubs) %>%
  dplyr::filter(
    is.finite(lon),
    is.finite(lat),
    is.finite(bsm_weight_pubs)
  )
```

# 2. Helper functions

```{r helpers}
clip_at_sd <- function(x, multiple = 1) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE) * multiple
  x %>%
    pmin(m + s) %>%
    pmax(m - s)
}

template_raster <- function(df) {
  template_df <- df %>%
    dplyr::select(lon, lat, gridid)
  sp::coordinates(template_df) <- c("lon", "lat")
  sp::gridded(template_df) <- TRUE
  sp::proj4string(template_df) <- sp::CRS(
    "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
  )
  raster::raster(template_df)
}
```

# 3. Apply the figure scaling

The paper states that Figure 3 maps were created using standard deviation scaling. Because the `bsm_weight_pubs` surface is strictly positive, this reproduction anchors the lower bound at `0` and uses `mean + 2.5 SD` as the upper color limit so the visual contrast better matches the published figure.

```{r scale-values}
eid_mean <- mean(predictions$bsm_weight_pubs, na.rm = TRUE)
eid_sd <- sd(predictions$bsm_weight_pubs, na.rm = TRUE)
eid_lower <- 0
eid_upper <- eid_mean + 2.5 * eid_sd

predictions$bsm_weight_pubs_clipped <- clip_at_sd(predictions$bsm_weight_pubs, 2.5)
```

# 4. Build the raster surface

```{r build-raster}
country_outlines <- terra::vect(
  sf::st_transform(
    rnaturalearth::ne_countries(scale = "medium", returnclass = "sf"),
    crs = "EPSG:4326"
  )
)

fig3a_raster <- predictions %>%
  dplyr::select(x = lon, y = lat, z = bsm_weight_pubs_clipped) %>%
  raster::rasterFromXYZ(crs = raster::crs(template_raster(predictions))) %>%
  raster::disaggregate(2, method = "bilinear")

fig3a_raster <- terra::rast(fig3a_raster)
terra::crs(fig3a_raster) <- "EPSG:4326"
fig3a_raster <- terra::mask(fig3a_raster, country_outlines)
names(fig3a_raster) <- "risk"

fig3a_df <- as.data.frame(fig3a_raster, xy = TRUE) %>%
  stats::na.omit()

world_df <- ggplot2::map_data("world")
```

# 5. Plot Figure 3A

```{r plot-fig3a}
fig_3a_plot <- ggplot() +
  geom_polygon(
    aes(x = long, y = lat, group = group),
    data = world_df,
    inherit.aes = FALSE,
    fill = viridis::viridis(1)
  ) +
  geom_tile(
    aes(x = x, y = y, fill = risk),
    data = fig3a_df
  ) +
  geom_path(
    aes(x = long, y = lat, group = group),
    data = world_df,
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.1
  ) +
  coord_fixed(xlim = c(-180, 180), ylim = c(-65, 90), expand = FALSE) +
  scale_fill_gradientn(
    colours = viridis::viridis(256),
    limits = c(eid_lower, eid_upper),
    oob = scales::squish,
    guide = guide_colorbar(
      label = TRUE,
      label.position = "right",
      title = "Observed EID risk index\n(0 to +2.5 SD)"
    )
  ) +
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    line = element_blank(),
    legend.title = element_text(color = "white", size = 8),
    legend.text = element_text(color = "white", size = 8),
    legend.title.align = 0,
    legend.background = element_blank(),
    legend.position = c(0.11, 0.45)
  ) +
  labs(
    title = "Figure 3A reproduction",
    subtitle = "Weighted model output with current reporting effort",
    x = NULL,
    y = NULL
  )

fig_3a_plot
```

# 6. Notes

- Variable used: `bsm_weight_pubs`
- Interpretation: predicted observed EID event risk under current reporting effort
- Color scale: anchored at `0` and clipped at `mean + 2.5 SD` to better match the published Figure 3A appearance

```{r optional-save, eval=FALSE}
ggsave("6.6.fig_3a_infectious_diseases.png", fig_3a_plot, width = 11, height = 6.5, dpi = 300)
```
