---
title: "Reproducing Figure 3A from Allen et al. (2017)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

if (!requireNamespace("librarian", quietly = TRUE)) {
  install.packages("librarian", repos = "https://cloud.r-project.org")
}

library(librarian)
shelf(raster, terra, sf, sp, rnaturalearth, tidyverse, viridis, maps)
```

# Overview

This document reproduces Figure 3A from Allen et al. (2017), *Global hotspots and correlates of emerging zoonotic diseases*.

Figure 3A is the predicted distribution of observed zoonotic EID event risk, using the weighted model output with current reporting effort.

Source:
[Allen et al. (2017), Nature Communications](https://www.nature.com/articles/s41467-017-00923-8)

# 1. Load archived model predictions

```{r load-predictions}
predictions_path <- "~/Library/CloudStorage/OneDrive-StockholmUniversity/KVA backup/KVAOneDrive_backup_28Jan2026/2. Projects/4.Cascades/Editorial Review 20250807/map/Base maps/6.Infectious_diseases/Allen 2017/predictions.RData"

stopifnot(file.exists(path.expand(predictions_path)))
load(path.expand(predictions_path))  # loads object: predictions

predictions <- predictions %>%
  dplyr::select(gridid, lon, lat, bsm_weight_pubs) %>%
  dplyr::filter(
    is.finite(lon),
    is.finite(lat),
    is.finite(bsm_weight_pubs)
  )
```

# 2. Helper functions

```{r helpers}
clip_at_sd <- function(x, multiple = 1) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE) * multiple
  x %>%
    pmin(m + s) %>%
    pmax(m - s)
}

template_raster <- function(df) {
  template_df <- df %>%
    dplyr::select(lon, lat, gridid)
  sp::coordinates(template_df) <- c("lon", "lat")
  sp::gridded(template_df) <- TRUE
  sp::proj4string(template_df) <- sp::CRS(
    "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
  )
  raster::raster(template_df)
}
```

# 3. Apply the figure scaling

The paper states that Figure 3 maps were created using standard deviation scaling. Because the `bsm_weight_pubs` surface is strictly positive, this reproduction anchors the lower bound at `0` and uses `mean + 2.5 SD` as the upper color limit so the visual contrast better matches the published figure.

```{r scale-values}
eid_mean <- mean(predictions$bsm_weight_pubs, na.rm = TRUE)
eid_sd <- sd(predictions$bsm_weight_pubs, na.rm = TRUE)
eid_lower <- 0
eid_upper <- eid_mean + 2.5 * eid_sd

predictions$bsm_weight_pubs_clipped <- clip_at_sd(predictions$bsm_weight_pubs, 2.5)
```

# 4. Build the raster surface

```{r build-raster}
country_outlines <- terra::vect(
  sf::st_transform(
    rnaturalearth::ne_countries(scale = "medium", returnclass = "sf"),
    crs = "EPSG:4326"
  )
)

fig3a_raster <- predictions %>%
  dplyr::select(x = lon, y = lat, z = bsm_weight_pubs_clipped) %>%
  raster::rasterFromXYZ(crs = raster::crs(template_raster(predictions))) %>%
  raster::disaggregate(2, method = "bilinear")

fig3a_raster <- terra::rast(fig3a_raster)
terra::crs(fig3a_raster) <- "EPSG:4326"
fig3a_raster <- terra::mask(fig3a_raster, country_outlines)
names(fig3a_raster) <- "risk"

fig3a_df <- as.data.frame(fig3a_raster, xy = TRUE) %>%
  stats::na.omit()

world_df <- ggplot2::map_data("world")
```

# 5. Plot Figure 3A

```{r plot-fig3a}
fig_3a_plot <- ggplot() +
  geom_polygon(
    aes(x = long, y = lat, group = group),
    data = world_df,
    inherit.aes = FALSE,
    fill = viridis::viridis(1)
  ) +
  geom_tile(
    aes(x = x, y = y, fill = risk),
    data = fig3a_df
  ) +
  geom_path(
    aes(x = long, y = lat, group = group),
    data = world_df,
    inherit.aes = FALSE,
    color = "white",
    linewidth = 0.1
  ) +
  coord_fixed(xlim = c(-180, 180), ylim = c(-65, 90), expand = FALSE) +
  scale_fill_gradientn(
    colours = viridis::viridis(256),
    limits = c(eid_lower, eid_upper),
    oob = scales::squish,
    guide = guide_colorbar(
      label = TRUE,
      label.position = "right",
      title = "Observed EID risk index\n(0 to +2.5 SD)"
    )
  ) +
  theme(
    panel.background = element_rect(fill = "black"),
    plot.background = element_rect(fill = "black"),
    line = element_blank(),
    legend.title = element_text(color = "white", size = 8),
    legend.text = element_text(color = "white", size = 8),
    legend.title.align = 0,
    legend.background = element_blank(),
    legend.position = c(0.11, 0.45)
  ) +
  labs(
    title = "Figure 3A reproduction",
    subtitle = "Weighted model output with current reporting effort",
    x = NULL,
    y = NULL
  )

fig_3a_plot
```

# 6. Comparison Overlay with the Summed Binary Pressure Surface

This section uses the same overlay approach as the comparison workflow in `6.5.comparison_maps.R`. The summed binary surface is loaded from the precomputed `summed_binary_context.rds` file, and the top `5%` of the Figure 3A risk surface is overlaid in black.

```{r comparison-overlay}
sum_context_path <- "/Users/egepehlivanoglu/Library/CloudStorage/OneDrive-StockholmUniversity/KVA backup/KVAOneDrive_backup_28Jan2026/2. Projects/4.Cascades/Editorial Review 20250807/map/Anthropocene-maps/summed_binary_context.rds"

stopifnot(file.exists(sum_context_path))
sum_context <- readRDS(sum_context_path)

target_crs <- sum_context$target_crs
target_res <- sum_context$target_res

rebuild_extent_from_df <- function(df, res_xy) {
  terra::ext(
    min(df$x, na.rm = TRUE) - res_xy[1] / 2,
    max(df$x, na.rm = TRUE) + res_xy[1] / 2,
    min(df$y, na.rm = TRUE) - res_xy[2] / 2,
    max(df$y, na.rm = TRUE) + res_xy[2] / 2
  )
}

binary_sum_df <- sum_context$binary_sum_df %>%
  dplyr::mutate(sum_binary_cat = factor(sum_binary, levels = c(0, 1, 2, 3, 4)))

target_extent <- if (!is.null(sum_context$target_extent)) {
  do.call(terra::ext, as.list(sum_context$target_extent))
} else {
  rebuild_extent_from_df(binary_sum_df, target_res)
}

world_moll <- if (!is.null(sum_context$world_moll)) {
  sum_context$world_moll
} else {
  sf::st_transform(
    rnaturalearth::ne_countries(scale = "medium", returnclass = "sf"),
    crs = target_crs
  )
}

common_template_raster <- terra::rast(
  ext = target_extent,
  resolution = target_res,
  crs = target_crs
)

fig3a_hist_df <- fig3a_df %>%
  dplyr::filter(!is.na(risk) & risk > 0)

fig3a_top5_cutoff <- stats::quantile(fig3a_hist_df$risk, probs = 0.95, na.rm = TRUE)
fig3a_top5_df <- fig3a_hist_df %>%
  dplyr::filter(risk >= fig3a_top5_cutoff)

fig3a_top5_vect <- terra::vect(fig3a_top5_df, geom = c("x", "y"), crs = "EPSG:4326")
fig3a_top5_vect_moll <- terra::project(fig3a_top5_vect, target_crs)
fig3a_top5_raster <- terra::rasterize(
  fig3a_top5_vect_moll,
  common_template_raster,
  field = "risk",
  fun = "max"
)

fig3a_top5_presence <- terra::ifel(is.na(fig3a_top5_raster), NA, 1)
names(fig3a_top5_presence) <- "presence"
fig3a_top5_presence_df <- as.data.frame(fig3a_top5_presence, xy = TRUE, na.rm = TRUE)

fig3a_overlay_plot <- ggplot() +
  geom_tile(
    data = binary_sum_df,
    aes(x = x, y = y, fill = sum_binary_cat)
  ) +
  geom_tile(
    data = fig3a_top5_presence_df,
    aes(x = x, y = y),
    fill = "yellow",
    alpha = 0.28
  ) +
  geom_sf(data = world_moll, fill = NA, color ="grey10", linewidth = 0.05) +
  coord_sf(crs = st_crs(target_crs), expand = FALSE) +
  scale_fill_manual(
    values = c(
      "0" = "#f7f7f7",
      "1" = "#d9e6f2",
      "2" = "#92b4d6",
      "3" = "#477db3",
      "4" = "#123b6d"
    ),
    na.value = "white",
    name = "Summed\nbinary layers"
  ) +
  labs(
    title = "Comparison Overlay: Figure 3A and Summed Binary Pressure",
    subtitle = "Black overlay marks the top 5% of Figure 3A infectious disease risk",
    caption = "The Figure 3A top 5% subset is calculated after excluding 0s and NAs from the Figure 3A risk surface.",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

fig3a_overlay_plot
```

# 7. Save Figure 3A and Figure 3B overlays side by side

- Variable used: `bsm_weight_pubs`
- Interpretation: predicted observed EID event risk under current reporting effort
- Color scale: anchored at `0` and clipped at `mean + 2.5 SD` to better match the published Figure 3A appearance
- Comparison overlay: uses the top `5%` of the Figure 3A risk surface after excluding `0s` and `NAs`

```{r save-both-overlays}
fig3b_env <- new.env(parent = globalenv())
sys.source("6.5.comparison_maps.R", envir = fig3b_env)
fig3b_overlay_plot <- get("fig3B_overlay_plot", envir = fig3b_env)

fig3a_overlay_export <- fig3a_overlay_plot +
  labs(
    title = "A. Figure 3A overlay",
    subtitle = "Top 5% of observed EID risk over summed binary pressure (weighted model output with current reporting effort)"
  )

fig3b_overlay_export <- fig3b_overlay_plot +
  labs(
    title = "B. Figure 3B overlay",
    subtitle = "Top 5% of bias-corrected EID risk over summed binary pressure (weighted model output reweighted by population)"
  )

pdf("6.6.fig3A_fig3B_overlay_comparison.pdf", width = 16, height = 6.5)
grid::grid.newpage()
grid::pushViewport(
  grid::viewport(layout = grid::grid.layout(nrow = 1, ncol = 2))
)
print(fig3a_overlay_export, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(fig3b_overlay_export, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))
dev.off()

png(
  "6.6.fig3A_fig3B_overlay_comparison.png",
  width = 16,
  height = 6.5,
  units = "in",
  res = 300
)
grid::grid.newpage()
grid::pushViewport(
  grid::viewport(layout = grid::grid.layout(nrow = 1, ncol = 2))
)
print(fig3a_overlay_export, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(fig3b_overlay_export, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))
dev.off()
```

# 8. Polytroubles Comparison for Figure 3A and Figure 3B

This comparison keeps the same yellow infectious overlays, but only colors summed-binary classes `2` and `3`. Classes `0`, `1`, and `NA` are shown alike.

```{r polytroubles-comparison}
fig3b_env <- new.env(parent = globalenv())
sys.source("6.5.comparison_maps.R", envir = fig3b_env)

fig3b_top5_presence_df <- get("top5_presence_df", envir = fig3b_env)
fig3b_world_moll <- get("world_moll", envir = fig3b_env)
fig3b_target_crs <- get("target_crs", envir = fig3b_env)
fig3b_binary_sum_df <- get("binary_sum_df", envir = fig3b_env)

make_polytroubles_plot <- function(sum_df, overlay_df, world_sf, crs_string, title_text, subtitle_text) {
  poly_df <- sum_df %>%
    dplyr::mutate(
      poly_class = dplyr::case_when(
        sum_binary == 2 ~ "2",
        sum_binary == 3 ~ "3",
        TRUE ~ NA_character_
      ),
      poly_class = factor(poly_class, levels = c("2", "3"))
    )

  ggplot() +
    geom_tile(
      data = poly_df,
      aes(x = x, y = y, fill = poly_class)
    ) +
    geom_tile(
      data = overlay_df,
      aes(x = x, y = y),
      fill = "yellow",
      alpha = 0.28
    ) +
    geom_sf(data = world_sf, fill = NA, color = "grey10", linewidth = 0.05) +
    coord_sf(crs = st_crs(crs_string), expand = FALSE) +
    scale_fill_manual(
      values = c(
        "2" = "#92b4d6",
        "3" = "#477db3"
      ),
      na.value = "white",
      name = "Summed\nbinary layers"
    ) +
    labs(
      title = title_text,
      subtitle = subtitle_text,
      caption = "Only summed-binary classes 2 and 3 are colored; 0, 1, and NA are shown alike.",
      x = NULL,
      y = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      axis.text = element_blank(),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold")
    )
}

fig3a_polytroubles_plot <- make_polytroubles_plot(
  sum_df = binary_sum_df,
  overlay_df = fig3a_top5_presence_df,
  world_sf = world_moll,
  crs_string = target_crs,
  title_text = "A. Figure 3A polytroubles",
  subtitle_text = "Top 5% of observed EID risk over summed-binary classes 2 and 3"
)

fig3b_polytroubles_plot <- make_polytroubles_plot(
  sum_df = fig3b_binary_sum_df,
  overlay_df = fig3b_top5_presence_df,
  world_sf = fig3b_world_moll,
  crs_string = fig3b_target_crs,
  title_text = "B. Figure 3B polytroubles",
  subtitle_text = "Top 5% of bias-corrected EID risk over summed-binary classes 2 and 3"
)

grid::grid.newpage()
grid::pushViewport(
  grid::viewport(layout = grid::grid.layout(nrow = 1, ncol = 2))
)
print(fig3a_polytroubles_plot, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(fig3b_polytroubles_plot, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))

pdf("6.6.fig3A_fig3B_polytroubles_comparison.pdf", width = 16, height = 6.5)
grid::grid.newpage()
grid::pushViewport(
  grid::viewport(layout = grid::grid.layout(nrow = 1, ncol = 2))
)
print(fig3a_polytroubles_plot, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(fig3b_polytroubles_plot, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))
dev.off()
```
