---
title: "Figure 3B Polytroubles With Yellow and Green Classes"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
project_root <- "/Users/egepehlivanoglu/Library/CloudStorage/OneDrive-StockholmUniversity/KVA backup/KVAOneDrive_backup_28Jan2026/2. Projects/4.Cascades/Editorial Review 20250807/map/Anthropocene-maps"
knitr::opts_knit$set(root.dir = project_root)

if (!requireNamespace("librarian", quietly = TRUE)) {
  install.packages("librarian", repos = "https://cloud.r-project.org")
}

library(librarian)
librarian::shelf(
  terra,
  sf,
  ggplot2,
  dplyr,
  tidyr,
  rnaturalearth,
  knitr
)
```

# Purpose

This notebook rebuilds the Figure 3B polytroubles map directly, using the same
classification logic used for the Figure 3A polytroubles map.

The key classes are:

- `eid`: Figure 3B top 5% EID risk only, shown in yellow
- `sum2_eid`: summed pressure class 2 plus Figure 3B top 5% EID risk, shown in light green
- `sum3_eid`: summed pressure class 3 plus Figure 3B top 5% EID risk, shown in dark green

The plot is rebuilt from data, rather than reusing the old `fig3B_polytroubles_plot`
object from `6.5`.

# 1. Load Summed Binary Pressure Context

```{r load-summed-pressure}
sum_context_path <- file.path(project_root, "summed_binary_context.rds")
stopifnot(file.exists(sum_context_path))

sum_context <- readRDS(sum_context_path)

target_crs <- sum_context$target_crs
target_res <- sum_context$target_res

binary_sum_df <- sum_context$binary_sum_df %>%
  dplyr::mutate(
    sum_binary = as.numeric(sum_binary),
    sum_binary_cat = factor(sum_binary, levels = c(0, 1, 2, 3, 4))
  )

rebuild_extent_from_df <- function(df, res_xy) {
  terra::ext(
    min(df$x, na.rm = TRUE) - res_xy[1] / 2,
    max(df$x, na.rm = TRUE) + res_xy[1] / 2,
    min(df$y, na.rm = TRUE) - res_xy[2] / 2,
    max(df$y, na.rm = TRUE) + res_xy[2] / 2
  )
}

target_extent <- if (!is.null(sum_context$target_extent)) {
  do.call(terra::ext, as.list(sum_context$target_extent))
} else {
  rebuild_extent_from_df(binary_sum_df, target_res)
}

common_template_raster <- terra::rast(
  ext = target_extent,
  resolution = target_res,
  crs = target_crs
)

world_moll <- if (!is.null(sum_context$world_moll)) {
  sum_context$world_moll
} else {
  sf::st_transform(
    rnaturalearth::ne_countries(scale = "medium", returnclass = "sf"),
    crs = target_crs
  )
}
```

# 2. Reproduce Figure 3B Risk Surface

```{r source-fig3b}
fig3b_script <- file.path(
  project_root,
  "6.Infectious_diseases",
  "6.2.Alan_et.al_fig3B_reproduced.R"
)

stopifnot(file.exists(fig3b_script))

fig3b_env <- new.env(parent = globalenv())
sys.source(fig3b_script, envir = fig3b_env)

fig3b_df <- get("fig3b_df", envir = fig3b_env)
```

# 3. Convert Figure 3B Top 5% Risk to the Common Grid

```{r fig3b-top5-presence}
fig3b_hist_df <- fig3b_df %>%
  dplyr::filter(!is.na(risk), risk > 0)

fig3b_top5_cutoff <- stats::quantile(
  fig3b_hist_df$risk,
  probs = 0.95,
  na.rm = TRUE
)

fig3b_top5_df <- fig3b_hist_df %>%
  dplyr::filter(risk >= fig3b_top5_cutoff)

fig3b_top5_vect <- terra::vect(
  fig3b_top5_df,
  geom = c("x", "y"),
  crs = "EPSG:4326"
)

fig3b_top5_vect_moll <- terra::project(fig3b_top5_vect, target_crs)

fig3b_top5_raster <- terra::rasterize(
  fig3b_top5_vect_moll,
  common_template_raster,
  field = "risk",
  fun = "max"
)
names(fig3b_top5_raster) <- "risk"

fig3b_top5_presence <- terra::ifel(is.na(fig3b_top5_raster), NA, 1)
names(fig3b_top5_presence) <- "presence"

fig3b_top5_presence_df <- as.data.frame(
  fig3b_top5_presence,
  xy = TRUE,
  na.rm = TRUE
)

fig3b_top5_cutoff
```

# 4. Build Fig 3B Polytroubles Classes

```{r build-polytroubles}
highlight_pressure_values <- c(2, 3)

fig3b_overlay_class_df <- fig3b_top5_presence_df %>%
  dplyr::transmute(x, y, infectious_presence = 1L)

fig3B_polytroubles_df <- binary_sum_df %>%
  dplyr::select(x, y, sum_binary) %>%
  dplyr::left_join(fig3b_overlay_class_df, by = c("x", "y")) %>%
  dplyr::mutate(
    infectious_presence = tidyr::replace_na(infectious_presence, 0L),
    poly_class = dplyr::case_when(
      sum_binary == 2 & infectious_presence == 0L ~ "sum2",
      sum_binary == 3 & infectious_presence == 0L ~ "sum3",
      infectious_presence == 1L &
        (is.na(sum_binary) | !(sum_binary %in% highlight_pressure_values)) ~ "eid",
      sum_binary == 2 & infectious_presence == 1L ~ "sum2_eid",
      sum_binary == 3 & infectious_presence == 1L ~ "sum3_eid",
      TRUE ~ NA_character_
    ),
    poly_class = factor(
      poly_class,
      levels = c("sum2", "sum3", "eid", "sum2_eid", "sum3_eid")
    )
  )

fig3B_polytroubles_counts <- fig3B_polytroubles_df %>%
  dplyr::count(poly_class, .drop = FALSE)

fig3B_polytroubles_counts

eid_only_n <- fig3B_polytroubles_counts$n[
  fig3B_polytroubles_counts$poly_class == "eid"
]

if (length(eid_only_n) == 1 && eid_only_n == 0) {
  message(
    "No yellow cells are present under the Fig 3A-style definition: ",
    "all Figure 3B top 5% EID cells overlap summed-pressure class 2 or 3."
  )
}
```

# 5. Plot Figure 3B Polytroubles

```{r plot-polytroubles, fig.width=11, fig.height=6.5}
polytroubles_palette <- c(
  "sum2" = "#9ecae1",
  "sum3" = "#3182bd",
  "eid" = "#ffd54f",
  "sum2_eid" = "#78c679",
  "sum3_eid" = "#238443"
)

polytroubles_labels <- c(
  "sum2" = "Summed pressure 2 only",
  "sum3" = "Summed pressure 3 only",
  "eid" = "Top 5% EID only",
  "sum2_eid" = "Overlap: 2 + EID",
  "sum3_eid" = "Overlap: 3 + EID"
)

fig3B_polytroubles_plot <- ggplot() +
  geom_tile(
    data = fig3B_polytroubles_df,
    aes(x = x, y = y, fill = poly_class)
  ) +
  geom_sf(data = world_moll, fill = NA, color = "grey10", linewidth = 0.05) +
  coord_sf(crs = sf::st_crs(target_crs), expand = FALSE) +
  scale_fill_manual(
    values = polytroubles_palette,
    labels = polytroubles_labels,
    drop = FALSE,
    na.value = "white",
    name = "Map class"
  ) +
  labs(
    title = "Figure 3B polytroubles",
    subtitle = "Top 5% bias-corrected EID risk, summed pressure, and overlap shown as separate classes",
    caption = "Yellow shows Figure 3B top 5% EID cells outside summed-pressure classes 2 and 3. Green shows overlap with summed-pressure classes 2 or 3.",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

fig3B_polytroubles_plot
```

# 6. Save Outputs

```{r save-output}
ggplot2::ggsave(
  filename = file.path("6.Infectious_diseases", "6.7.fig3B_polytroubles_yellow_green.png"),
  plot = fig3B_polytroubles_plot,
  width = 11,
  height = 6.5,
  dpi = 300
)

ggplot2::ggsave(
  filename = file.path("6.Infectious_diseases", "6.7.fig3B_polytroubles_yellow_green.pdf"),
  plot = fig3B_polytroubles_plot,
  width = 11,
  height = 6.5
)
```
