---
title: "Raster Math Dataset Summary"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

project_root <- "/Users/egepehlivanoglu/Library/CloudStorage/OneDrive-StockholmUniversity/KVA backup/KVAOneDrive_backup_28Jan2026/2. Projects/4.Cascades/Editorial Review 20250807/map/Anthropocene-maps"
knitr::opts_knit$set(root.dir = project_root)

summarise_object <- function(name, env) {
  obj <- get(name, envir = env)
  cat("###", name, "\n\n")
  cat("- Class:", paste(class(obj), collapse = ", "), "\n")
  if (inherits(obj, "SpatRaster")) {
    cat("- Layers:", terra::nlyr(obj), "\n")
    cat("- CRS:", terra::crs(obj), "\n")
    cat("- Resolution:", paste(terra::res(obj), collapse = " x "), "\n")
    cat("- Extent:", paste(as.vector(terra::ext(obj)), collapse = ", "), "\n")
    cat("- Names:", paste(names(obj), collapse = ", "), "\n")
  } else if (inherits(obj, "sf")) {
    cat("- Rows:", nrow(obj), "\n")
    cat("- CRS:", sf::st_crs(obj)$input, "\n")
    cat("- Geometry type(s):", paste(unique(as.character(sf::st_geometry_type(obj))), collapse = ", "), "\n")
  } else {
    cat("- Structure:\n")
    print(utils::str(obj))
  }
  cat("\n")
}

source_and_keep <- function(script_path, keep_names) {
  env <- new.env(parent = globalenv())
  sys.source(script_path, envir = env)
  rm(list = setdiff(ls(envir = env), keep_names), envir = env)
  env
}
```

# Overview

This document sources the latest raster-math scripts from:

- `1.Deforestation`
- `2.Conflict`
- `4.Climate_change`
- `5.Pesticide_use`

For each script, only the last two dataset objects defined at the end of the file are retained and summarised.

# 1. Deforestation

**Source file:** `1.Deforestation/1.4.Deforestation_bin.R`

```{r deforestation-source}
deforestation_env <- source_and_keep(
  "1.Deforestation/1.4.Deforestation_bin.R",
  c("deforestation_bin_SR", "world_moll_deforestation_SR")
)
```

```{r deforestation-summary, results='asis'}
summarise_object("deforestation_bin_SR", deforestation_env)
summarise_object("world_moll_deforestation_SR", deforestation_env)
```

# 2. Conflict

**Source file:** `2.Conflict/2.5.Conflict_bin.R`

```{r conflict-source}
conflict_env <- source_and_keep(
  "2.Conflict/2.5.Conflict_bin.R",
  c("conflict_bin_SR", "conflict_land_sf_SR")
)
```

```{r conflict-summary, results='asis'}
summarise_object("conflict_bin_SR", conflict_env)
summarise_object("conflict_land_sf_SR", conflict_env)
```

# 4. Climate Change

**Source file:** `4.Climate_change/4.4.Climate_change_bin_SR.R`

```{r climate-source}
climate_env <- source_and_keep(
  "4.Climate_change/4.4.Climate_change_bin_SR.R",
  c("clim_change_binSR", "world_moll_clim")
)
```

```{r climate-summary, results='asis'}
summarise_object("clim_change_binSR", climate_env)
summarise_object("world_moll_clim", climate_env)
```

# 5. Pesticide Use

**Source file:** `5.Pesticide_use/5.3.Pesticide_RiskScores_SRbin.R`

```{r pesticide-source}
pesticide_env <- source_and_keep(
  "5.Pesticide_use/5.3.Pesticide_RiskScores_SRbin.R",
  c("pesticide_SR_bin", "pesticide_world_moll")
)
```

```{r pesticide-summary, results='asis'}
summarise_object("pesticide_SR_bin", pesticide_env)
summarise_object("pesticide_world_moll", pesticide_env)
```
