---
title: "Summed Binary SpatRaster Map"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

project_root <- "/Users/egepehlivanoglu/Library/CloudStorage/OneDrive-StockholmUniversity/KVA backup/KVAOneDrive_backup_28Jan2026/2. Projects/4.Cascades/Editorial Review 20250807/map/Anthropocene-maps"
knitr::opts_knit$set(root.dir = project_root)

library(terra)
library(sf)
library(ggplot2)
library(dplyr)
library(rnaturalearth)

source_and_keep <- function(script_path, keep_names) {
  env <- new.env(parent = globalenv())
  sys.source(script_path, envir = env)
  rm(list = setdiff(ls(envir = env), keep_names), envir = env)
  env
}

binary_to_zero <- function(r) {
  ifel(is.na(r), 0, r)
}
```

# Purpose

This document sources the latest binary raster workflows and sums the binary `SpatRaster` outputs into one composite map.

Because raster math requires aligned grids, all rasters are first reprojected/resampled to a common Mollweide template. To avoid creating false precision, the common template uses the **coarsest resolution** among the input rasters.

# 1. Source the binary rasters

```{r source-rasters}
deforestation_env <- source_and_keep(
  "1.Deforestation/1.4.Deforestation_bin.R",
  c("deforestation_bin_SR", "world_moll_deforestation_SR")
)

conflict_env <- source_and_keep(
  "2.Conflict/2.5.Conflict_bin.R",
  c("conflict_bin_SR", "conflict_land_sf_SR")
)

climate_env <- source_and_keep(
  "4.Climate_change/4.4.Climate_change_bin_SR.R",
  c("clim_change_binSR", "world_moll_clim")
)

pesticide_env <- source_and_keep(
  "5.Pesticide_use/5.3.Pesticide_RiskScores_SRbin.R",
  c("pesticide_SR_bin", "pesticide_world_moll")
)

r_list <- list(
  deforestation = get("deforestation_bin_SR", envir = deforestation_env),
  conflict = get("conflict_bin_SR", envir = conflict_env),
  climate = get("clim_change_binSR", envir = climate_env),
  pesticide = get("pesticide_SR_bin", envir = pesticide_env)
)
```

# 2. Check CRS and resolution

```{r inspect-geometry}
raster_geometry <- tibble(
  dataset = names(r_list),
  crs = vapply(r_list, terra::crs, character(1)),
  res_x = vapply(r_list, function(r) terra::res(r)[1], numeric(1)),
  res_y = vapply(r_list, function(r) terra::res(r)[2], numeric(1)),
  xmin = vapply(r_list, function(r) terra::ext(r)[1], numeric(1)),
  xmax = vapply(r_list, function(r) terra::ext(r)[2], numeric(1)),
  ymin = vapply(r_list, function(r) terra::ext(r)[3], numeric(1)),
  ymax = vapply(r_list, function(r) terra::ext(r)[4], numeric(1))
)

raster_geometry
```

# 3. Create a common raster template

```{r common-template}
target_crs <- terra::crs(r_list$deforestation)
target_res <- c(
  max(vapply(r_list, function(r) terra::res(r)[1], numeric(1))),
  max(vapply(r_list, function(r) terra::res(r)[2], numeric(1)))
)

target_extent <- terra::ext(
  min(vapply(r_list, function(r) terra::ext(r)[1], numeric(1))),
  max(vapply(r_list, function(r) terra::ext(r)[2], numeric(1))),
  min(vapply(r_list, function(r) terra::ext(r)[3], numeric(1))),
  max(vapply(r_list, function(r) terra::ext(r)[4], numeric(1)))
)

template_raster <- terra::rast(
  ext = target_extent,
  resolution = target_res,
  crs = target_crs
)

template_raster
```

# 4. Align rasters before raster math

```{r align-rasters}
r_aligned <- lapply(r_list, function(r) {
  terra::resample(r, template_raster, method = "near")
})

r_aligned_zero <- lapply(r_aligned, binary_to_zero)
```

# 5. Sum the binary rasters

```{r sum-rasters}
binary_sum <- Reduce(`+`, r_aligned_zero)
names(binary_sum) <- "sum_binary"

binary_sum_df <- as.data.frame(binary_sum, xy = TRUE, na.rm = FALSE)
```

# 6. Plot in Mollweide projection

```{r plot-sum}
world_moll <- sf::st_transform(
  rnaturalearth::ne_countries(scale = "medium", returnclass = "sf"),
  crs = target_crs
)

binary_sum_plot <- ggplot() +
  geom_tile(
    data = binary_sum_df,
    aes(x = x, y = y, fill = sum_binary)
  ) +
  geom_sf(
    data = world_moll,
    fill = NA,
    color = "black",
    linewidth = 0.1
  ) +
  coord_sf(crs = st_crs(target_crs), expand = FALSE) +
  scale_fill_viridis_c(
    option = "C",
    na.value = "white",
    name = "Count of\nbinary layers"
  ) +
  labs(
    title = "Summed Binary Raster Math Surface",
    subtitle = "Deforestation + Conflict + Climate Change + Pesticide Use",
    caption = "All binary rasters aligned to a common Mollweide template using the coarsest input resolution before summation.",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

binary_sum_plot
```

# 7. Final output objects

```{r final-objects}
binary_sum
target_res
```
