---
title: "Deforestation binary"
output: html_notebook
---

# Workflow
# 1. Load data (GeoTIFF from Zenodo)
# 2. Investigate data (resolution, CRS, extent, dimensions)
> r
class       : SpatRaster 
size        : 14000, 36000, 1  (nrow, ncol, nlyr)
resolution  : 0.01, 0.01  (x, y)
extent      : -180, 180, -56, 84  (xmin, xmax, ymin, ymax)
coord. ref. : lon/lat WGS 84 (EPSG:4326) 
source      : spat_80851fac04a7_32901_P2mm2aa9P9QX7pk.tif 
varname     : file808546a31610 
name        : file808546a31610_1 
min value   :                  1 
max value   :                  7 

# 3. Change Resolution Reproject to Mollweide
Resoulution is changed because otherwise I receive this watning.
Warning: Too many points (529 out of 529) failed to transform, unable to compute output bounds. (GDAL error 1)

> r_moll
class       : SpatRaster 
size        : 1526, 3608, 1  (nrow, ncol, nlyr)
resolution  : 10000, 10000  (x, y)
extent      : -18040096, 18039904, -6486272, 8773728  (xmin, xmax, ymin, ymax)
coord. ref. : +proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs 
source(s)   : memory
name        : file808546a31610_1 
min value   :                  1 
max value   :                  7 

# 4. Binary transformation (permanent vs temporary) 

# 5. Plot with ggplot


# Load Packages
```{r}
# install librarian package (working better than pacman)
# From CRAN:
install.packages("librarian")
librarian::shelf(httr, terra, tmap, tidyverse) # install and load packages
```

# 1. Load Data
```{r}
# 1) Download the public GeoTIFF (v1.2, 2001–2024) from Zenodo
url <- "https://zenodo.org/records/15366671/files/drivers_forest_loss_1km_2001_2024_v1_2.tif?download=1"  # Zenodo direct
tif_path <- tempfile(fileext = ".tif")
res <- GET(url, write_disk(tif_path, overwrite = TRUE))
stop_for_status(res)
```

# 2. Investigate data

```{r Arrange Data and Inspect}
# 2) Read the raster
# Band 1 = dominant class (uint8, 1..7); Bands 2..8 = class probabilities (0..250)
r_all <- rast(tif_path)
r <- r_all[[1]]          # dominant class only
r <- clamp(r, 1, 7,  values = FALSE)      # be safe: anything outside 1..7 to NA
# FALSE is added recently
# Investigate data
# r is your SpatRaster object. Here are some key properties to check:
r
# Resolution (pixel size in map units)
res(r)
# CRS / projection (WKT string)
crs(r)
# A friendlier summary: EPSG if recognized + proj string
terra::crs(r, describe=TRUE)
# Extent (bounding box) in CRS units
ext(r)
# Dimensions: nrows, ncols, nlyr
dim(r)
# Number of cells
ncell(r)
# convert data to categorical (factor) for plotting
r <- as.factor(r)        # categorical
```

# 3. Change Resolution

Resolution changed to 10 km

```{r}
# 3) Reproject to Mollweide (categorical -> nearest neighbor)
#r_moll <- project(r, "ESRI:54009", method = "near")
#use new projection since many points failed to transform (specify the resolution)
moll <- "+proj=moll +lon_0=0 +datum=WGS84 +units=m +no_defs"
r_moll <- terra::project(r, moll, method="near", res=10000)
r_moll
```

# 4. Binary transformation for high resolution

```{r}
# metadata: -1 = non-ag land, -2 = water/no data  -> set to NA
r_moll[r_moll %in% c(-1, -2)] <- NA              # keep raw counts for everything else
names(r_moll) <- "value"

# # ---- target CRS: Mollweide ----
# # # Using a standard proj string for world Mollweide
#  crs_moll <- "+proj=moll +lon_0=0 +datum=WGS84 +units=m +no_defs"
# # 
# # # ---- reproject raster (use nearest neighbor to preserve raw counts) ----
#  r_moll2 <- project(r_moll, crs_moll, method = "near")

# (Optional) speed up plotting if huge:
# r_moll <- aggregate(r_moll, fact = 2, fun = "first")

# ---- country boundaries -> sf -> Mollweide ----
world <- ne_countries(scale = "medium", returnclass = "sf")
world_moll <- st_transform(world, crs = crs_moll)
gc() #important so the memory limit is OK
# ---- to data frame for ggplot ----
df <- as.data.frame(r_moll, xy = TRUE, na.rm = TRUE)  # columns: x, y, value
# Attach human-readable labels from the dataset docs (v1.2)
# Codes: 1 Perm. agriculture, 2 Hard commodities, 3 Shifting cultivation,
# 4 Logging, 5 Wildfire, 6 Settlements & infrastructure, 7 Other natural disturbances
lev <- data.frame(
  ID   = 1:7,
  name = c("Permanent agriculture",
           "Hard commodities",
           "Shifting cultivation",
           "Logging",
           "Wildfire",
           "Settlements & infrastructure",
           "Other natural disturbances"),
  binary_codes =c(1,1,1,0,0,0,0)
)

df_bin <- merge(df, lev, by.x = "value", by.y = "ID", all.x = TRUE)  %>% # set value as factorial and merge with lev to get name and binary codes
# convert value name, binary_codes columns to factor
  mutate(name = factor(name, levels = lev$name), 
         binary_codes = factor(binary_codes, levels = c(0,1))) %>% 
  select(x, y, value, name, binary_codes) # select only the columns we need
```


# 5. Plot
```{r}
bin_plot_hr <- ggplot() +
  geom_tile(data = df_bin, aes(x = x, y = y, fill = binary_codes)) +
  geom_sf(data = world_moll, fill = NA, color = "black", linewidth = 0.2) +
  coord_sf() +
  scale_fill_manual(
    name = "Disturbance Types",
    values = c("0" = "#f4d03f",  # low risk (example yellow)
               "1" = "#c0392b",  # high risk (example red)
               "NA" = "white"),  # explicitly map NA to white
    na.value = "white",           # ensures NAs are white
    labels = c("0" = "Temporary", "1" = "Permanent")
  ) +
  labs(
    title = "Deforestation Plot",
    subtitle = "Binary Transformed Data (10000 m)",
    caption = "Permanent Disturbances:Permanent agriculture, Hard commodities, Shifting cultivation. Temporary Disturbances: Logging, Wildfire, Settlements & infrastructure, and Other natural disturbances",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    # panel.background = element_rect(fill = "gray85", color = NA), # gray background (ocean)
    # plot.background = element_rect(fill = "gray85", color = NA),
    panel.grid = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.key.height = unit(0.6, "cm"),
    legend.text = element_text(size = 9)
  )
bin_plot_hr

#save plot
# pdf(file = "2026-02-17_deforestation_binary_plot.pdf", width = 11, height = 6.5)
# bin_plot_hr
# dev.off()
```


