---
title: "Deforestation binary"
output:
  html_document:
    df_print: paged
---

## Workflow

1. **Load data** (GeoTIFF from Zenodo)
2. **Inspect the raster** (resolution, CRS, extent, dimensions)
3. **Reproject to Mollweide** (after adjusting resolution)
4. **Binary transformation** (permanent vs temporary)
5. **Plot** (ggplot)

---

## 1–2. Data import and quick inspection

### Raster summary

> **Object:** `r`  
> **Type:** `SpatRaster` (terra)

| Field | Value |
|---|---|
| Class | `SpatRaster` |
| Size (nrow, ncol, nlyr) | **14000**, **36000**, **1** |
| Resolution (x, y) | **0.01**, **0.01** |
| Extent (xmin, xmax, ymin, ymax) | **-180**, **180**, **-56**, **84** |
| CRS | lon/lat WGS 84 (**EPSG:4326**) |
| Source | `spat_80851fac04a7_32901_P2mm2aa9P9QX7pk.tif` |
| Varname | `file808546a31610` |
| Name | `file808546a31610_1` |
| Min–Max | **1** – **7** |

---

## 3. Reproject to Mollweide (after changing resolution)

::: {.callout-warning}
### Why change the resolution first?

Reprojecting directly at **0.01°** caused GDAL to fail to compute output bounds:

> *Warning: Too many points (529 out of 529) failed to transform, unable to compute output bounds. (GDAL error 1)*

To avoid this, the raster is first aggregated / resampled to a coarser grid before reprojection.
:::

<details>
<summary><strong>Mollweide raster summary (`r_moll`)</strong></summary>

| Field | Value |
|---|---|
| Class | `SpatRaster` |
| Size (nrow, ncol, nlyr) | **1526**, **3608**, **1** |
| Resolution (x, y) | **10000**, **10000** |
| Extent (xmin, xmax, ymin, ymax) | **-18040096**, **18039904**, **-6486272**, **8773728** |
| CRS | `+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs` |
| Source | `memory` |
| Name | `file808546a31610_1` |
| Min–Max | **1** – **7** |

</details>

# Load Packages
```{r Load Packages}
# install librarian package (working better than pacman)
# From CRAN:
install.packages("librarian", repos = "https://cloud.r-project.org")
library(librarian)
shelf(httr, terra, tmap, tidyverse, rnaturalearth, sf) # install and load packages
```

# 1. Load Data
```{r 1. Load Data}
# 1) Download the public GeoTIFF (v1.2, 2001–2024) from Zenodo
url <- "https://zenodo.org/records/15366671/files/drivers_forest_loss_1km_2001_2024_v1_2.tif?download=1"  # Zenodo direct
tif_path <- tempfile(fileext = ".tif")
res <- GET(url, write_disk(tif_path, overwrite = TRUE))
stop_for_status(res)
```

# 2. Investigate data

```{r 2. Investigate data}
# 2) Read the raster
# Band 1 = dominant class (uint8, 1..7); Bands 2..8 = class probabilities (0..250)
r_all <- rast(tif_path)
r <- r_all[[1]]          # dominant class only
r <- clamp(r, 1, 7,  values = FALSE)      # be safe: anything outside 1..7 to NA
# FALSE is added recently
# Investigate data
# r is your SpatRaster object. Here are some key properties to check:
r
# Resolution (pixel size in map units)
res(r)
# CRS / projection (WKT string)
crs(r)
# A friendlier summary: EPSG if recognized + proj string
terra::crs(r, describe=TRUE)
# Extent (bounding box) in CRS units
ext(r)
# Dimensions: nrows, ncols, nlyr
dim(r)
# Number of cells
ncell(r)
# convert data to categorical (factor) for plotting
r <- as.factor(r)        # categorical
```

# 3. Change Resolution

Resolution changed to 10 km

```{r 3. Change Resolution}
# 3) Reproject to Mollweide (categorical -> nearest neighbor)
#r_moll <- project(r, "ESRI:54009", method = "near")
#use new projection since many points failed to transform (specify the resolution)
moll <- "+proj=moll +lon_0=0 +datum=WGS84 +units=m +no_defs"
r_moll <- terra::project(r, moll, method="near", res=10000)
r_moll
```

# 4. Binary transformation for high resolution

```{r 4. Binary transformation for high resolution}
# metadata: -1 = non-ag land, -2 = water/no data  -> set to NA
r_moll[r_moll %in% c(-1, -2)] <- NA              # keep raw counts for everything else
names(r_moll) <- "value"
# ---- target CRS: Mollweide ----
# Using a standard proj string for world Mollweide
crs_moll <- "+proj=moll +lon_0=0 +datum=WGS84 +units=m +no_defs"

# ---- country boundaries -> sf -> Mollweide ----
world <- ne_countries(scale = "medium", returnclass = "sf")
world_moll <- st_transform(world, crs = crs_moll)
gc() #important so the memory limit is OK
# ---- to data frame for ggplot ----
df <- as.data.frame(r_moll, xy = TRUE, na.rm = TRUE)  # columns: x, y, value
# Attach human-readable labels from the dataset docs (v1.2)
# Codes: 1 Perm. agriculture, 2 Hard commodities, 3 Shifting cultivation,
# 4 Logging, 5 Wildfire, 6 Settlements & infrastructure, 7 Other natural disturbances
lev <- data.frame(
  ID   = 1:7,
  name = c("Permanent agriculture",
           "Hard commodities",
           "Shifting cultivation",
           "Logging",
           "Wildfire",
           "Settlements & infrastructure",
           "Other natural disturbances"),
  binary_codes =c(1,1,1,0,0,0,0)
)

df_bin <- merge(df, lev, by.x = "value", by.y = "ID", all.x = TRUE)  %>% # set value as factorial and merge with lev to get name and binary codes
  mutate(name = factor(name, levels = lev$name), # convert name, binary_codes columns to factor
         binary_codes = factor(binary_codes, levels = c(0,1)))
deforestation_bin <- df_bin
```


# 5. Plot
```{r 5. Plot}
bin_plot_hr <- ggplot() +
  geom_tile(data = df_bin, aes(x = x, y = y, fill = binary_codes)) +
  geom_sf(data = world_moll, fill = NA, color = "black", linewidth = 0.1) +
  coord_sf() +
  scale_fill_manual(
    name = "Disturbance Types",
    values = c("0" = "seagreen4",  # low risk (example yellow)
               "1" = "salmon4",  # high risk (example red)
               "NA" = "white"),  # explicitly map NA to white
    na.value = "white",           # ensures NAs are white
    labels = c("0" = "Temporary", "1" = "Permanent")
  ) +
  labs(
    title = "Deforestation Plot",
    subtitle = "Binary Transformed Data (10000 m)",
    caption = "Permanent Disturbances:Permanent agriculture, Hard commodities, Shifting cultivation. Temporary Disturbances: Logging, Wildfire, Settlements & infrastructure, and Other natural disturbances",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    # panel.background = element_rect(fill = "gray85", color = NA), # gray background (ocean)
    # plot.background = element_rect(fill = "gray85", color = NA),
    panel.grid = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.key.height = unit(0.6, "cm"),
    legend.text = element_text(size = 9)
  )
bin_plot_hr

# #save plot
# pdf(file = "2026-02-17_deforestation_binary_plot.pdf", width = 11, height = 6.5)
# bin_plot_hr
# dev.off()
```


# Optional 
If you want to extract R chunks automatically to .R script use the code below. 

```{r create a script}
#knitr::purl("1.Deforestation/1.2.Deforestation_binary.Rmd")
```

